---
title: "Phenotype residualization"
output: html_document
---

Input: round 2's ancestry-filtered keep-list (person IDs) and a phenotype list TSV. Output: one `FID IID Y` file per phenotype x transform x covariate-set combination, matching `GRM-pairs/full_grm_bin/prep_pheno.R`'s expected format.

Protocol (order matters, see Kemper et al. 2021 Methods):

1. residualize `phenotype ~ covariates`
2. trim residuals > 5 SD from the mean
3. standardize surviving residuals to mean 0, var 1, within each sex

Covariates: sex-at-birth + age always included. PCs / zip3 (factor) / SES vars (`zip3_ses_map`) are each independently toggleable -- every combination gets run, not just one chosen set.

## Setup

**Run this chunk first in a freshly restarted R session.** If a package (e.g. `rlang`) already got auto-loaded from the system library before this chunk runs -- which can happen just from session startup -- `.libPaths()` can't retroactively swap it for the newer version installed here; R won't unload/replace an already-attached namespace mid-session. Symptom: `namespace 'rlang' 1.1.6 is already loaded, but >= 1.1.7 is required`. Fix is a full session restart (Session > Restart R in RStudio), not re-running chunks in the same session.

```{r setup}
# R_LIBS_USER is set on this environment, but to a colon-separated list of
# non-writable system paths (conda + spark), not a single writable directory --
# install.packages(lib=...) needs one directory, so don't trust R_LIBS_USER here.
# Always use an explicit personal library instead.
user_lib <- file.path(Sys.getenv("HOME"), "R", "library")
dir.create(user_lib, recursive = TRUE, showWarnings = FALSE)
.libPaths(c(user_lib, .libPaths()))

required_pkgs <- c("dplyr", "tidyr", "readr", "stringr", "e1071", "bigrquery")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) install.packages(missing_pkgs, lib = user_lib)

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(bigrquery)
source("../../scripts/local/residualize_lib.R")   # transform, residualize, export -- shared with the fake-data test
```

## Inputs

- `KEEP_LIST_PATH`: round 2 output -- person IDs passing ancestry filtering.
- `PHENO_LIST_PATH`: TSV describing which phenotypes to pull (see schema below).
- `PC_PATH`: round 1's PCA output (`.eigenvec`), or round 1's AoU-projected `.sscore` once that step is done.
- `ZIP3_SES_TABLE`: AoU's `zip3_ses_map` -- exact join path (likely via `observation`, not directly on `person`) still needs confirming against the real workbench, flagged below.

Phenotype list TSV schema (`PHENO_LIST_PATH`):

| column | purpose |
|---|---|
| `phenotype_name` | label used in output filenames/diagnostics |
| `source` | `measurement` / `survey` / `condition` -- picks the extraction path |
| `concept_id` | AoU concept ID (comma-separated for multiple) |
| `value_field` | optional -- which field holds the numeric value |
| `notes` | free text, not used programmatically |

```{r inputs}
KEEP_LIST_PATH <- "PLACEHOLDER"   # round 2 output, one person_id per line
PHENO_LIST_PATH <- "../../docs/phenotype_list.tsv"  # starter anthropometric/metabolic panel -- see its header for concept_id confirmation status
PC_PATH <- "PLACEHOLDER"          # round 1 .eigenvec (or AoU-projected .sscore)
OUT_DIR <- "PLACEHOLDER"          # where the FID/IID/Y files get written

dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
```

## Read the ID list and phenotype list

`keep_ids`: everyone downstream gets restricted to this set before anything else -- residualization, transforms, and diagnostics are all computed within this cohort only, not the full AoU population.

```{r read-lists}
keep_ids <- read_lines(KEEP_LIST_PATH)

pheno_list <- read_tsv(PHENO_LIST_PATH, col_types = cols(.default = "c"))
stopifnot(all(c("phenotype_name", "source", "concept_id") %in% names(pheno_list)))
pheno_list
```

## Pull phenotype + covariate table

**Unresolved, needs the real workbench to pin down:**

- exact BigQuery table/column names per `source` type (`measurement` vs `survey` vs `condition`)
- `zip3_ses_map` join path
- age definition -- age at the specific measurement/response (more correct, varies per phenotype) vs. a single fixed age pulled once. Written here as per-phenotype since that's the more defensible default; simplify later if it turns out not to matter.

Everything below this chunk (transform, covariate-set configs, residualization, export, diagnostics) doesn't depend on these specifics and is fully generic once `pull_phenotype()` returns a table shaped `person_id, phenotype, age, sex_at_birth`.

```{r pull-functions}
pull_phenotype <- function(row, keep_ids) {
  # TODO: real BigQuery query, branching on row$source ("measurement" / "survey" / "condition").
  # Must return: person_id, phenotype (numeric), age (numeric, at the phenotype's own
  # measurement/response time), sex_at_birth (factor) -- restricted to keep_ids.
  stop("pull_phenotype() not implemented -- fill in against the real workbench")
}

pull_covariates <- function(keep_ids) {
  pcs <- read_table(PC_PATH, col_types = cols(.default = "d", IID = "c")) %>%
    rename(person_id = IID) %>%
    filter(person_id %in% keep_ids)

  zip3 <- tibble(person_id = character(), zip3 = character())        # TODO: real join
  ses <- tibble(person_id = character(), median_income = double())    # TODO: real join

  list(pcs = pcs, zip3 = zip3, ses = ses)
}
```

## Optional skew-reducing transform

Rank-based inverse-normal transform (`inverse_normal_transform()` in `residualize_lib.R`) -- chosen over log/Box-Cox since it doesn't assume a particular skew direction or require positive values, so it works uniformly across an arbitrary phenotype list without per-phenotype tuning. Both `raw` and `invnorm` variants get carried through the rest of the pipeline; skewness before/after is reported so it's clear whether the transform actually helped for each phenotype, rather than applying it blindly. See `02_phenotype/notebooks/local/test_residualize_fake_data.ipynb` (or its eventual `.Rmd` counterpart) for a worked example on synthetic data showing this diagnostic actually catching a real skew reduction.

## Covariate-set configs

`build_covariate_sets()` (in `residualize_lib.R`) returns a named list of formula RHS vectors. `sex_at_birth` is handled separately (it's the stratification variable for step 3, not a residualization covariate -- see `residualize_phenotype()` in the lib), so it's not in these formulas.

```{r covariate-sets}
pc_cols <- paste0("PC", 1:20)   # matches round 1's 20-PC output
covariate_sets <- build_covariate_sets(pc_cols)
names(covariate_sets)
```

## Main loop

`run_residualization()` (in `residualize_lib.R`) runs the full cross product of {phenotype} x {raw, invnorm} x {covariate-set}, all exported -- no curation, every combination in `pheno_list` x `covariate_sets` gets written out as an `FID IID Y` file matching `GRM-pairs/full_grm_bin/prep_pheno.R`'s expected format.

```{r main-loop}
result <- run_residualization(
  pheno_list, keep_ids, pull_phenotype, pull_covariates,
  covariate_sets, OUT_DIR
)
```

## Diagnostics summary

```{r diagnostics-skew}
result$skew_summary_table   # per phenotype: skewness before/after the invnorm transform
```

```{r diagnostics-combo}
result$combo_summary_table  # per phenotype x variant x covariate-set: N retained, R^2
```
